A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-22 sections/Discussion.tex
S2.4 scoping; hat alpha format

If the reader is not familiar with DK's algorithm,
he/she might be confused about the inconsistency across judgment.
As an example, $(\lam x)~()$ synthesizes $\al$,
but the second premise of the subsumption rule uses $1$ for the result.
This is because a context application $[\Gm,\al=1,\bt=\al]\al = 1$ happens between the premises.

The existential variables $\al$ and $\bt$ are clearly not used after the subsumption rule,
but according to the algorithm, they are kept in the context
until some parent judgment recycles a block of variables,
or to the very end of a type inference task.
In that sense, DK's algorithm does not control the scoping of variables precisely enough.
However, it is a minor issue that does not affect the soundness and completeness properties.

>>>>

If the reader is not familiar with DK's algorithm,
he/she might be confused about the inconsistent types across judgment.
As an example, $(\lam x)~()$ synthesizes $\al$,
but the second premise of the subsumption rule uses $1$ for the result.
This is because a context application $[\Gm,\al=1,\bt=\al]\al = 1$ happens between the premises.

The existential variables $\al$ and $\bt$ are clearly not used after the subsumption rule,
but according to the algorithm, they are kept in the context
until some parent judgment recycles a block of variables,
or to the very end of a type inference task.
In that sense, DK's algorithm does not control the scoping of variables precisely enough.
%However, it is a minor issue that does not affect the soundness and completeness properties.

### 2019-02-22 sections/Overview.tex
S2.4 scoping; hat alpha format

\jimmy{Possibly some discussions about the typing rules}

@@

DK's algorithm version revolves around their notion of \emph{algorithmic context}.
\[
\begin{array}{l@{\qquad}lcl}
\text{Algorithmic Contexts}\qquad&\Gamma,\Delta,\Theta &::=&\quad \nil \mid
\Gamma, a \mid \Gamma, \hat{\alpha} \mid \Gamma, \hat{\alpha} = \tau \mid
\Gamma, \blacktriangleright_{\hat{\alpha}}
\end{array}
\]
In addition to the regular (universally quantified) type variables $a$, the
algorithmic context also contains \emph{existential} type variables
$\hat{\alpha}$. These are placeholders for monotypes $\tau$ that are still to
be determined by the inference algorithm. When the existential variable is
``solved'', its entry in the context is replaced by the assignment
$\hat{\alpha} = \tau$. With the notation $[\Gamma]$ an algorithmic context
$\Gamma$ denotes the substitution that replaces all its solved existential type
variables with their solution.

@@

\paragraph{Markers for scoping.}
The sequential order of entries in the algoritmic context, in combination with
the threading of contexts,  does not perfectly capture the scoping of all
existential variables. For this reason the DK algorithm uses scope markers
$\blacktriangleright_{\hat{\alpha}}$ in a few places. An example is given in the following
rule:
\[
\inferrule*
  {\Gamma,\blacktriangleright_{\hat{\alpha}}, \hat{\alpha} \vdash [\hat{\alpha}/a]A \le B \dashv \Delta,\blacktriangleright_{\hat{\alpha}},\Theta}
  {\Gamma \vdash \forall a. A \le B \dashv \Delta}
\]
To indicate that the scope of $\hat{\alpha}$ is local to the subtyping check
$[\hat{\alpha}/a]A \le B$, the marker is pushed onto its input stack and popped
from the output stack together with the subsequent part $\Theta$, which may
refer to $\hat{\alpha}$. \jimmy{which may contain definition/solution of $\al$ and references to $\al$}
(Remember that later entries may refer to earlier
ones, but not vice versa.) This way $\hat{\alpha}$ does not escape its scope.

\jimmy{Do we talk about scoping control without markers as well here?\\
e.g. $\forall I$ ($e \Lto \all A$) or $\le\forall R$ ($A \le \all B$)\\
Text about garbage refers to this point!}

Due to the complicated scoping control method,
it is hard to argue the preciseness of scoping for each variable when designing algorithms.
We found a scoping issue related to the subsumption rule,
which might cause existential variable leakage across judgments.
A detailed discussion on this issue with a possible fix is in Section~\ref{sec:discussion:scoping}.

>>>>

@@

DK's algorithm version revolves around their notion of \emph{algorithmic context}.
\[
\begin{array}{l@{\qquad}lcl}
\text{Algorithmic Contexts}\qquad&\Gamma,\Delta,\Theta &::=&\quad \nil \mid
\Gamma, a \mid \Gamma, \al \mid \Gamma, \al = \tau \mid
\Gamma, \blacktriangleright_{\al}
\end{array}
\]
In addition to the regular (universally quantified) type variables $a$, the
algorithmic context also contains \emph{existential} type variables
$\al$. These are placeholders for monotypes $\tau$ that are still to
be determined by the inference algorithm. When the existential variable is
``solved'', its entry in the context is replaced by the assignment
$\al = \tau$. With the notation $[\Gamma]$ an algorithmic context
$\Gamma$ denotes the substitution that replaces all its solved existential type
variables with their solution.

@@

\paragraph{Markers for scoping.}
The sequential order of entries in the algoritmic context, in combination with
the threading of contexts,  does not perfectly capture the scoping of all
existential variables. For this reason the DK algorithm uses scope markers
$\blacktriangleright_{\al}$ in a few places. An example is given in the following
rule:
\[
\inferrule*
  {\Gamma,\blacktriangleright_{\al}, \al \vdash [\al/a]A \le B \dashv \Delta,\blacktriangleright_{\al},\Theta}
  {\Gamma \vdash \forall a. A \le B \dashv \Delta}
\]
To indicate that the scope of $\al$ is local to the subtyping check
$[\al/a]A \le B$, the marker is pushed onto its input stack and popped
from the output stack together with the subsequent part $\Theta$, which may
refer to $\al$. \jimmy{which may contain definition/solution of $\al$ and references to $\al$}
(Remember that later entries may refer to earlier
ones, but not vice versa.) This way $\al$ does not escape its scope.

A type variable may have a similar functionality to the scoping markers.
An example rule that checks an expression against a polymorphic type is as follows:
$$
\inferrule*
{\Gm, a \vdash e \Lto A \dashv \Delta, a, \Theta}
{\Gm \vdash e \Lto \all A \dashv \Delta}
$$

Due to the complication of its scoping control method,
it is hard to argue the preciseness of scoping for each variable when designing algorithms.
Actually we found a scoping issue related to the subsumption rule,
which might cause existential variable leakage across judgments.
A detailed discussion on this issue with a possible fix is in Section~\ref{sec:discussion:scoping}.
